---
title: How it's been doing a statistics PhD plus some research
subtitle: EA Imperial
author: Adam Howes
institute: Imperial College London
date: February 2022
bibliography: citations.bib
output:
  beamer_presentation:
    latex_engine: pdflatex
    highlight: haddock
    fig_width: 7 
    fig_height: 3
    includes:
        in_header: preamble.tex
---

## Academic background

* Maths undergraduate at Durham, no real ideas about career
  * Worked in a pub first year, Cranfield second year, and Basque Center for Applied Mathematics third year
  * Dissertation about Bayesian optimisation 
* Not enough statistics courses at Durham, so did a masters at Warwick
  * Dissertation about Bayesian models for combining sources of evidence
* StatML CDT (Imperial / Oxford) started the year I was applying
  * Somewhat successor to OxWaSP (Oxford / Warwick)

## EA background

* Heard about 80,000 Hours in 3rd year from a friend
* Listened to a few podcasts, found them compelling, first time anything career related really resonated with me
* Read a few books over summer, then started going to the EA society at Warwick
  * Events like retreats, more people in the community, helped to add depth
* First EAG I went to was the year I started at Imperial
  * Volunteering can be a good way to get a place

## Any tips about applying for PhDs

* Find (and call if possible) any previous students of the supervisor
  * What are they doing now?
* Assess the vibe among current graduate students
* Be interested in things, read papers, attend reading groups
  * "List of things you don't need permission to do"
* Doing research as an undergraduate is a good test

## Reflections on PhD experience

* Many (all?) of my friends doing PhDs have had periods where it has negatively effected them
  * "The well-being costs of graduate school versus STIs"
* Little structure, often sporadic feedback, uncertainty about research direction and progress, making a narrow contribution (that perhaps few people care about) to a large field
* Different to undergraduate: not learning material then passing test for another 3/4 years

## What's my PhD about

* HIV/AIDS is an infectious disease which is endemic in sub-Saharan Africa
* Epidemic indicators can be used to guide policy response
* My area of research is Bayesian spatio-temporal statistics, the method we use to perform small-area estimation
  * (I'll explain what these terms mean)

## Bayesian statistics

* First step in Bayesian statistics is to write a probabilistic model for the data
  * Focus is on understanding and modelling scientific phenomenon, rather than e.g. remembering arcane statistical tests
* Second step is to perform inference (which is usually done with an algorithm)
  * Given we observed certain data, what can we learn about the parameters of the model?
* Things might go wrong with the above two steps with the data, probabilistic model, algorithm
  * Should also perform checks, following a "Bayesian workflow"

## Spatio-temporal statistics

* Say that you have a collection of people who have taken a HIV test, and their results are either positive or negative
* If have the location for these people, this might be important information to include in our probabilistic model
* Similarly, the time that they took the test might also be important
* Spatio-temporal statistics is just when we think these two features of the data are important to take into account
* It allows us to answer spatio-temporal questions
  * What's the situation at time $t$ and location $s$?
  * Often important for policy

## My work

* I'm interested in applied and methodological work
  * Applied: focus is on answering a question about the world
  * Methodological: focus is on developing techniques which can be used to answer questions about the world
* The deliniation between these might not be clear sometimes
  * To answer questions about the world you'll often have to develop new methods
  * To develop new (relevant) methods that you should be thinking about which questions they will allow you to answer to think
* Some examples of my work
  * Applied: estimating numbers of young women at risk
  * Methodological: investigating models for spatial smoothing

## Applied example

* Women 15-29 are disproportionately at risk of HIV
* To efficiently target programs and resources to the most at risk women, we're working on small-area estimates of the number of women in different risk groups

##

\begin{table}
\centering
\begin{tabularx}{\textwidth}{lXlX}
\toprule
Category & Description & Risk ratio \\ 
 \midrule
None & Not sexually active & 0 \\
Low & One cohabiting sexual partner & 1 \\
High & Non-regular sexual partner(s) & 1.72  \\
Very high & Young women from key populations & 13 \\
\bottomrule
\end{tabularx}
\caption{HIV risk categories and risk ratios.}
\label{tab:risk-categories}
\end{table}

##

```{r echo=FALSE}
knitr::include_graphics("depends/category-flowchart.pdf")
```

##

```{r echo=FALSE}
knitr::include_graphics("depends/available-surveys.pdf")
```

##

```{r echo=FALSE}
knitr::include_graphics("depends/aids-abstract-A.pdf")
```

##

```{r echo=FALSE}
knitr::include_graphics("depends/aids-abstract-B.pdf")
```

## Methods example

* TO-DO: Create annotated equations version of the Besag model

## Career plans going forward

* Global health
* Biosecurity
  * Reading group this term, Fridays at St Mary's 12-1!

## Thanks!
